---
title: "R tutorial series: Analyzing bacterial growth from OD600 data"
subtitle: ""
summary: "This article will include code chunks on how I analyze bacterial growth of OD600 data"
authors: [admin]
tags: [Portfolio, r, dataviz]
categories: []
date: 2024-08-08T18:23:23+07:00
lastmod: 2024-08-08T18:23:23+07:00
featured: false
draft: false
output:
  blogdown::html_page:
    toc: false

image:
  caption: ""
  focal_point: ""
  preview_only: true

projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.retina = 2)
```

Bacterial growth analysis is a fundamental task in microbiology, allowing researchers to monitor the proliferation of bacteria over time. Optical Density at 600 nm (OD600) is a commonly used method to measure this growth. In this tutorial, we'll walk through the process of analyzing bacterial growth data using R: loading and tidying the data, visualizing growth curves for several samples at once, automatically detecting each sample's exponential phase to estimate its maximum specific growth rate (**&mu;max**), and comparing growth rates across samples. This is just basic things for a new undergrad biology student and a nice way to start learning microbiology.


# Libraries we need
We start by defining the libraries that will be used in our analysis.

```{r}
libs <- c(
  "readr", "dplyr", "tidyr", "purrr", "lubridate",
  "ggplot2", "ggsci", "plotly", "knitr")

installed_libs <- libs %in% rownames(installed.packages())
if (any(installed_libs == FALSE)) {
  install.packages(libs[!installed_libs])
}

invisible(lapply(libs, library, character.only = TRUE))
```

`readr` for reading the CSV, `dplyr`/`tidyr`/`purrr` for wrangling, `lubridate` for parsing the time stamps, `ggplot2`/`ggsci` for static plots, `plotly` for an interactive version, and `knitr` for printing tidy summary tables.

# Load and tidy the data
We load the OD600 data from a CSV file (Excel works too, just keep the same column layout). You can download example data [here](https://drive.google.com/file/d/1O6Y1g-_Z6oRyfYlfth24hm-s7GWBqKhF/view?usp=share_link). This is personal data collected during my training at the University of Vienna, using *Vibrio* as a bacterial model grown in marine broth liquid medium.

The raw file has one row per reading, with a `Sample_name` column encoding both the sample and the time it was taken (e.g. `WT_09:15`), plus the raw `Result` (OD600 reading). The workflow below supports any number of samples — nothing has to change if you add more.

```{r}
blank_od <- 0.001 # OD600 of your blank/media-only well — adjust to your own measurement

df <- read_csv("od_data2.csv") %>%
  mutate(
    Sample = sub("_.*", "", Sample_name),
    Time   = sub(".*_", "", Sample_name),
    Time   = hm(Time),
    Result = Result - blank_od) %>%
  filter(Result > 0) %>% # can't take log() of a non-positive OD
  mutate(Result_log = log(Result)) %>%
  group_by(Sample) %>%
  arrange(Time, .by_group = TRUE) %>%
  mutate(Time_h = as.numeric(Time - min(Time), units = "hours")) %>%
  ungroup() %>%
  select(Sample, Time_h, Result, Result_log)

glimpse(df)
```

A few changes from a "just read the CSV" approach worth calling out:

- The blank correction lives in a named `blank_od` variable instead of a magic number buried in a pipe, so it's obvious what to change for your own plate reader.
- Any blank-corrected reading `<= 0` is dropped before log-transforming — otherwise `log()` silently returns `-Inf`/`NaN` and breaks the regressions further down.
- `Time` is converted to **hours elapsed since that sample's first reading** (`Time_h`), computed per sample. That keeps the x-axis in intuitive units and means samples that were started at different clock times still line up at t = 0.

# Growth curves for all samples
With tidy data in hand, we can plot every sample's growth curve on one panel.

```{r}
p_raw <- ggplot(df, aes(x = Time_h, y = Result, color = Sample)) +
  geom_line() +
  geom_point(size = 1) +
  labs(
    title = "Optical Density λ600",
    x = "Time (hours)",
    y = "Optical Density (OD600)") +
  scale_color_nejm() +
  theme_minimal()

p_raw
```

# Log-transformed OD600
Plotting `log(OD600)` against time is what makes the exponential phase visible as a straight line — that straight segment is exactly what the next section detects automatically.

```{r}
p_log <- ggplot(df, aes(x = Time_h, y = Result_log, color = Sample)) +
  geom_line() +
  geom_point(size = 1) +
  labs(
    title = "Optical Density λ600 (log scale)",
    x = "Time (hours)",
    y = "log(Optical Density)") +
  scale_color_nejm() +
  theme_minimal()

p_log
```

# Finding &mu;max with a sliding-window regression

In batch culture, growth during the exponential phase follows

$$ \frac{dN}{dt} = \mu N $$

so a straight line fit to `log(N)` vs. time has slope &mu;, the specific growth rate. **&mu;max** is the highest &mu; sustained anywhere in the curve — i.e. the steepest straight stretch of the log-transformed curve.

Rather than eyeballing where the exponential phase starts and ends (as I did in an earlier version of this post, and which has to be re-guessed by hand for every new sample), we slide a small window of consecutive time points along each sample's curve, fit a linear regression of `Result_log ~ Time_h` inside every window, and keep the window with the steepest good-quality fit. This is the same idea used by dedicated growth-curve tools like the `growthcurver` package.

```{r}
find_mu_max <- function(data, window = 5) {
  data <- arrange(data, Time_h)
  n <- nrow(data)
  if (n < window) {
    return(tibble())
  }

  map_dfr(1:(n - window + 1), function(i) {
    w <- data[i:(i + window - 1), ]
    fit <- lm(Result_log ~ Time_h, data = w)
    tibble(
      window_start = min(w$Time_h),
      window_end   = max(w$Time_h),
      mu           = unname(coef(fit)[2]),
      r_squared    = summary(fit)$r.squared
    )
  })
}
```

`window` is the number of consecutive readings used per regression — tune it to your sampling frequency (e.g. a smaller window for coarser sampling, a larger one for very frequent readings) so each window still spans a few minutes of real growth.

We apply this per sample, then for each sample keep the best window: the steepest slope among windows with a good linear fit (`r_squared > 0.98`).

```{r}
growth_windows <- df %>%
  group_by(Sample) %>%
  group_modify(~ find_mu_max(.x, window = 5)) %>%
  ungroup()

mu_max_summary <- growth_windows %>%
  filter(r_squared > 0.98, mu > 0) %>%
  group_by(Sample) %>%
  slice_max(mu, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(doubling_time_min = (log(2) / mu) * 60) %>%
  select(Sample, mu_max = mu, r_squared, window_start, window_end, doubling_time_min)

kable(mu_max_summary, digits = 3,
      col.names = c("Sample", "μmax (h⁻¹)", "R²",
                    "Window start (h)", "Window end (h)", "Doubling time (min)"))
```

The doubling time follows directly from &mu;max:

$$ t_d = \frac{\ln(2)}{\mu_{max}} $$

# Checking the detected exponential phase
Before trusting these numbers, it's worth visualizing which window was picked for each sample, overlaid on the log-transformed curve.

```{r, fig.height=4}
ggplot(df, aes(x = Time_h, y = Result_log)) +
  geom_rect(
    data = mu_max_summary,
    aes(xmin = window_start, xmax = window_end, ymin = -Inf, ymax = Inf),
    inherit.aes = FALSE, fill = "grey80", alpha = 0.5
  ) +
  geom_line(aes(color = Sample), show.legend = FALSE) +
  geom_point(aes(color = Sample), size = 1, show.legend = FALSE) +
  facet_wrap(~Sample, scales = "free_x") +
  labs(
    title = "Detected exponential-phase window per sample",
    x = "Time (hours)", y = "log(Optical Density)") +
  scale_color_nejm() +
  theme_minimal()
```

The shaded rectangle is the window `find_mu_max()` selected for that sample. If it looks too short, too noisy, or off to one side, adjust the `window` size or the `r_squared` cutoff above and re-run.

# Comparing growth rate across samples
This is the part that's easy to lose track of when analyzing one sample at a time: how do the samples actually compare?

```{r, fig.height=4}
mu_max_summary %>%
  select(Sample, `μmax (h⁻¹)` = mu_max, `Doubling time (min)` = doubling_time_min) %>%
  pivot_longer(-Sample, names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = reorder(Sample, value), y = value, fill = Sample)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~metric, scales = "free_x") +
  coord_flip() +
  scale_fill_nejm() +
  labs(x = NULL, y = NULL, title = "Growth rate vs. doubling time by sample") +
  theme_minimal()
```

# Interactive growth curve
Handy for zooming into a specific stretch of the curve or checking individual points.

```{r}
ggplotly(p_raw)
```

# Remarks
By automating exponential-phase detection with a sliding-window regression, this workflow scales to any number of samples without hand-tuning a `lag_end`/`exp_end` cutoff for each one, and the &mu;max/doubling-time table and every plot above are regenerated straight from the raw CSV whenever this document is knitted. That makes the whole analysis reproducible: re-run it on a new plate and the numbers and figures update themselves.

Buy me a [coffee](https://bmc.link/danilyanedo)
See you in the next post and have a beautiful day!
